Staged Abstract Interpreters

### 2019-03-25 paper/20_prelim.tex
section 3

\section{Preliminaries}

@@

Now we can instantiate the interpreter as a standard environment-store
interpreter. Such interpreter can be obtained by refunctionalization of CESK
machines \cite{Felleisen:1987:CAH:41625.41654, DBLP:conf/ppdp/AgerBDM03}.

@@

\paragraph{Unstaged Concrete Monads}
% The data manipulated by the monad are staged value. But the monad is known at
% current stage.
The environment and store can be modeled by reader effect and state effect,
i.e., a Reader monad and a State monad. We combine them using monad transformers
@ReaderT@ and @StateT@.
In other word, we instantiate the @AnsM@ moand as a stack of @ReaderT@ and @StateT@
transformers, where the @ReaderT@ is parameterized by type @Env@, @StateT@ is
parameterized by type @Store@, and the inner-most monad @IdM@ is just an identity
monad.

@@

To apply a function, the method @ap_clo@ takes a function value and an arugment
value, extracts the lambda term and environment enclosed in the function value,
and evaluates the body expression of that lambda under the new environment and store.

>>>>

\section{Preliminaries} \label{prelim}

@@

Now we can instantiate the interpreter as a standard environment-store
interpreter. Such interpreter also can be obtained by refunctionalization of
CESK machines \cite{Felleisen:1987:CAH:41625.41654, DBLP:conf/ppdp/AgerBDM03}.

@@

\paragraph{Unstaged Monads}
The environment and store can be modeled by reader effect and state effect,
i.e., a Reader monad and a State monad. We combine them using monad transformers
@ReaderT@ and @StateT@.
In other word, we instantiate the @AnsM@ moand as a stack of @ReaderT@ and @StateT@
transformers\footnote{The question mark syntax is a kind projector
  \cite{kindprojector}, thus \texttt{StateT[IdM,Store,?]} is equivalent to \newline
  \texttt{(\{type M[T]=StateT[IdM,Store,T]\})\#M}}, where the @ReaderT@ is
parameterized by type @Env@, @StateT@ is parameterized by type @Store@, and the
inner-most monad @IdM@ is just an identity monad.

@@

To apply a function, the method @ap_clo@ takes a function value and an arugment
value, extracts the lambda term and environment enclosed in the function value,
and evaluates the body expression of that lambda term under the new environment and store.

### 2019-03-25 paper/30_stageconc.tex
section 3

\subsection{Multi-Stage Programming in LMS}

Lightweight modular staging (LMS) \cite{DBLP:conf/gpce/RompfO10} is a
multi-stage programming framework implemented as a Scala library that enables
dynamic code generation in a type-safe manner. Different from the syntactic
approach in MetaOCaml \cite{DBLP:conf/flops/Kiselyov14,
DBLP:conf/gpce/CalcagnoTHL03} that uses quotations, LMS distinguishes
binding-time solely based on types. LMS provides a type constructor @Rep[T]@
where @T@ can be an arbitrary type, indicating an expression will be known at
next stage. All operations acting on a @Rep[T]@ expression will be residualized
in the generated code. However, we need to provide intermediate representation
and code generation support for type @T@ to LMS. Fortunately, the LMS framework
already provides such support for primitive types and commonly used data
structure such as arrays and maps; implementing such support for custom class is
also straightforward. Let's go back and see how LMS can be used to specialize
the power function we mentioned in Section 1.

@@

In the code shown above, \texttt{power} takes two arguments where \texttt{b} is
declared as \texttt{Rep[Int]} type meaning that \texttt{b} is a representation
of \texttt{Int} but whose value will be available at the next stage. Meanwhile,
the result of the @power@ function will be available at the next stage as well,
thus the return value of \texttt{power} is also of type \texttt{Rep[Int]}. Then
we specialize @power@ in function @snippet@ by providing @x@ to @5@. The
generated code for \texttt{power5} is a function that takes only one argument
which corresponds to @b@, and the body of this function multiplies @b@ five
times.

@@

@@

\begin{lstlisting}
trait RepConcInterpOps extends Concrete with LMSOps {
  type R[+T] = Rep[T]
  val $\rho$0: Rep[Env] = Map[Ident,Addr]()
  val $\sigma$0: Rep[Store] = Map[Addr,Value]()
  def get($\rho$: Rep[Env], x: Ident): Rep[Addr] = $\rho$(x)
  def put($\rho$: Rep[Env], x: Ident, a: Rep[Addr]): Rep[Env] = 
    $\rho$ + (unit(x) -> a)
  def get($\sigma$: Rep[Store], a: Rep[Addr]): Rep[Value] = $\sigma$(a)
  def put($\sigma$: Rep[Store], a: Rep[Addr], v: Rep[Value]): Rep[Store] = 
    $\sigma$ + (a -> v)
  def alloc($\sigma$: Rep[Store], x: Ident): Rep[Addr] = $\sigma$.size + 1
  def num(i: Lit): Rep[Value] = lift(NumV(i))
  def branch0(cnd: Rep[Value], thn: => Ans, els: => Ans): Ans = {
    //FIXME: the cast is ugly, patten match?
    val i = cnd.asInstanceOf[Rep[NumV]].i
    if (i == 0) thn else els
  }
  def prim_eval(op: Symbol, 
    v1: Rep[Value], v2: Rep[Value]): Rep[Value] = op match {
      case '+ => v1.asInstanceOf[Rep[NumV]].i + 
                 v2.asInstanceOf[Rep[NumV]].i
      ...
    }
  ...
}
\end{lstlisting}

The first notable change is that the abstract type member @R@ is assigned to be
@Rep@, and accordingly the affected types such as @Env@ and @Store@ become
@Rep[Env]@ and @Rep[Store]@. For @num@, we use the @lift@ function which is a
built-in in LMS that lifts a current-stage value to next stage. The interesting
point is how we handle closures and function applications when staging is
involved. In the unstaged version, function @close@ is used to lift a literal
lambda term to a closure, which is a pair of the syntactic lambda term and the
enclosing environment. However, what we desire is a \textit{compiled} closure,
instead of a lifted @CloV@ value like what we did for @NumV@ --- in other words,
the syntactic term of the lambda expression should be eliminated and specialized
away. The specialization of the interpreter with respect to the body of the
lambda term proceeds under the assumptions that the argument and latest store
will be provided later.
%\note{to denotational style?, back to HO?}
We will see this is an important observation that enables us to specialize an
abstract interpreter in a modular way. At the end of function @close@, we create
a @CompiledClo@ object that contains a staged function, which all will be
emitted in the generated code.

@@

\begin{lstlisting}
  case ApplyClosure(f, arg, $\sigma$) => 
    emitValDef(sym, quote(f) + 
                    ".asInstanceOf[CompiledClo].f(" + 
                    quote(arg) + "," + 
                    quote($\sigma$) + ")")
\end{lstlisting}

>>>>

In this section, based on the generic interpreter and concrete components we
presented in Section \ref{prelim}, we show how to obtain a staged concrete
interpreters by changing the monads and refactoring several primitive
operations. We begin by briefly introducing the Lightweight Modular Staging
framework in Scala.

\subsection{Multi-Stage Programming with LMS}

Lightweight modular staging (LMS) \cite{DBLP:conf/gpce/RompfO10} is a
multi-stage programming framework implemented as a Scala library that enables
dynamic code generation in a type-safe manner. Different from the approach of
MetaOCaml \cite{DBLP:conf/flops/Kiselyov14, DBLP:conf/gpce/CalcagnoTHL03} that
uses syntactic quotation and escape, LMS distinguishes binding-time solely based
on types. LMS provides a type constructor @Rep[T]@ where @T@ can be an arbitrary
type, indicating a value of @T@ will be known at next stage. All operations
acting a @Rep[T]@ expression will be residualized as generated code.

The LMS framework provides staging support for primitive data types such as
@Int@ and @Double@, and commonly-used data structures such as arrays and maps.
The idea is to construct intermediate representations (IR) for the next-stage
program at current stage. Such conversion from expressions of type @Rep[_]@ to
their IR is done by using implicit design pattern. As we will see later,
implementing the staging support for custom classes is also straightforward.

For example, let us go back and see how LMS can be used to specialize the
@power@ function from Section ~\ref{intro}.

@@

\todo{Show some IR}

@@

To implement the staged concrete interpreter, we reuse the exactly same concrete
components from Section \ref{unstaged_conc}, but we instantiate the binding-time
type @R[T]@ as @Rep[T]@.

\paragraph{Staged Monads} We also need to use monads that operate on staged
values. Note that the monadic objects such as @StateT@ or the internal value
@R[S] => M[(A,S)]@ are not staged, only the data @S@ or @A@ are staged.
This is why we can completely eliminate the overhead caused by the monadic layers.
Here we use the prefix name @Rep@ to differentitate unstaged monads and staged
monads, such as @RepReaderT@. So the @AnsM[_]@ is instantiated as the same monad
stack before, but with @Rep@ versions.

\begin{lstlisting}
  type R[T] = Rep[T]
  type AnsM[T] = RepReaderT[RepStateT[RepIdM, Store, ?], Env, T]
\end{lstlisting}

We use @RepStateT@ as an example to give a sketch of how the @Rep@ versions of monad
transformers look like. The generic type @M[_]@ is a @RepMonad@ and used as the
inner monad, which means the pair @(A,S)@ is also staged when appears inside of monad
@M[_]@. The @flatMap@ is similar to the unstaged one, but the function @f@ takes
a @Rep[A]@ value. Again, @f@ is not a next-stage function, but a current-stage
function that transforms a next-stage value of @Rep[A]@ to @StateT[M,S,B]@.
\todo{highlight the changes in the code}

\begin{lstlisting}
  case class RepStateT[M[_]: RepMonad, S, A](run: Rep[S] => M[(A, S)]) {
    def flatMap[B](f: Rep[A] => StateT[M, S, B]): StateT[M, S, B] = ...
  }
\end{lstlisting}

Readers may notice that the conversion between unstaged monads and monads is 
simply changing the unstaged data to staged data, which in fact can be achieved without
changing the implementation of monads. This is true so far, but as will see in
the abstract interpreter when introducing the nondeterminism monad (@ListT@), as
the whole list will be a next-stage value instead of merely the elements in the list.
For this reason, we explicitly distinguish and present the two version of
monads, as a more consistent approach.

\paragraph{Primitive Operations} Most of the primitive operations can be easily
translated to staged version -- we just need to let them works on staged data.

\begin{lstlisting}
  def emit_ap_clo(fun: Rep[Value], arg: Rep[Value], σ: Rep[Store]): Rep[(Value, Store)]
  def emit_compiled_clo(fun: (Rep[Value], Rep[Store]) => Rep[(Value, Store)],
                        λ: Lam, ρ: Rep[Env]): Rep[Value]

  def close(ev: EvalFun)(λ: Lam, ρ: Rep[Env]): Rep[Value] = {
    val Lam(x, e) = λ
    val f: (Rep[Value], Rep[Store]) => Rep[(Value, Store)] = {
      case (v: Rep[Value], σ: Rep[Store]) =>
        val α = alloc(σ, x)
        ev(e)(ρ + (unit(x) → α))(σ + (α → v))
    }; emit_compiled_clo(f, λ, ρ)
  }
  def ap_clo(ev: EvalFun)(fun: Rep[Value], arg: Rep[Value]): Ans = for {
    σ <- get_store
    val res: Rep[(Value, Store)] = emit_ap_clo(fun, arg, σ)
    _ <- put_store(res._2)
  } yield res._1
\end{lstlisting}

\begin{lstlisting}
  case class IRApClo(f: Exp[Value], arg: Exp[Value], σ: Exp[Store]) extends Def[(Value, Store)]
  override def emit_ap_clo(fun: Exp[Value], arg: Exp[Value], σ: Exp[Store]): Exp[(Value, Store)] =
    IRApClo(fun, arg, σ)
\end{lstlisting}

\begin{lstlisting}
  override def emitNode(sym: Sym[Any], rhs: Def[Any]) = rhs match {
    case IRApClo(f, arg, σ) =>
      emitValDef(sym, s"$\textdollar${quote(f)}.asInstanceOf[CompiledClo].f($\textdollar${quote(arg)}, $\textdollar${quote(σ)})")
    ...
  }
\end{lstlisting}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

The first notable change is that the abstract type member @R@ is assigned to be
@Rep@, and accordingly the affected types such as @Env@ and @Store@ become
@Rep[Env]@ and @Rep[Store]@. For @num@, we use the @lift@ function which is a
built-in in LMS that lifts a current-stage value to next stage. The interesting
point is how we handle closures and function applications when staging is
involved. In the unstaged version, function @close@ is used to lift a literal
lambda term to a closure, which is a pair of the syntactic lambda term and the
enclosing environment. However, what we desire is a \textit{compiled} closure,
instead of a lifted @CloV@ value like what we did for @NumV@ --- in other words,
the syntactic term of the lambda expression should be eliminated and specialized
away. The specialization of the interpreter with respect to the body of the
lambda term proceeds under the assumptions that the argument and latest store
will be provided later.

We will see this is an important observation that enables us to specialize an
abstract interpreter in a modular way. At the end of function @close@, we create
a @CompiledClo@ object that contains a staged function, which all will be
emitted in the generated code.

@@
